---
title: "EDA"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Read in the Bike Details Data

```{r}
library(tidyverse)
bikeData <- read_csv("https://www4.stat.ncsu.edu/~online/datasets/bikeDetails.csv")
bikeData %>% 
  select(selling_price, year, km_driven, owner, seller_type, everything())
```

Let's convert some `chr` to `factors` and consider the `NA` or missing values.

```{r}
#convert seller_type to a factor
bikeData <- bikeData %>% 
  mutate(seller_type_factor = as.factor(seller_type))
#find missing value numbers by column
colSums(is.na(bikeData))
#what to do with those missing values? problem dependent
```

## Graphs and Numeric Summaries

Start with summarizing the categorical variables (i.e. `str` or `factor` variables)

```{r}
table(bikeData$seller_type_factor)
```
What does this mean???

```{r}
table(bikeData$owner, bikeData$year)
```

Same thing in graphical form: bar graph
```{r}
ggplot(bikeData, aes(x = owner)) +
  geom_bar()
```


Now we'll summarize some of the numeric variables. `selling_price` is what we care about the most so focus on that!

```{r}
bikeData %>%
  group_by(owner, seller_type_factor) %>%
  summarize(mean = mean(selling_price),
            median = median(selling_price),
            sd = sd(selling_price),
            min = min(selling_price),
            max = max(selling_price))
```

Some graphs of `selling_price`. 

```{r}
ggplot(bikeData, aes(x = km_driven, y = selling_price))+
  geom_point()
```


Do a log transform on both!
```{r}
bikeData <- bikeData %>%
  mutate(log_selling_price = log(selling_price),
         log_km_driven = log(km_driven))
ggplot(bikeData, aes(x = log_km_driven, y = log_selling_price))+
  geom_point()
```





