---
title: "Creating graphs"
author: "Craig"
date: '2023-11-04'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("tidyverse")
```

## How to create graphs from the Stack Overflow survey

Each of the questions have been formatted so that R can easily read and manipulate the results. Each question is stored as it's own file at the following website.

<https://javastats.com/data/asa_data_challenge/>

To analyze the results of a single question we must complete the follwoing:

#1 Read in the data

Identify the question of interest. Right click on the hyperlink for that question and copy the link. We can then use read_csv to read in the file from the website.

```{r}
# Reading in the employment question results

employment <- read_csv("https://javastats.com/data/asa_data_challenge/survey_results_expand_on_column_Employment.csv")
full_data <- read_csv("https://www4.stat.ncsu.edu/~online/datasets/survey_results_public.csv")
```

When you examine the data you will see that has created a binary option for each response, 1 for present and 0 for not present. In order to summarize this data we need to find how many 1's and 0's are present for each column. 

#2 Summarize the data and store it in a data frame

First we can find the column sums. 
```{r}
summary_employ <-colSums(employment)

```

This creates a single vector with the column sums. We will want to attach the survey outcome to each of these numbers. Luckily, this vector is a named vector meaning the name information is present. We can extract it and save it as its own vector. 

```{r}
employment_names <- names(summary_employ)
```

Now we just need to put these two vectors together in a data frame.

```{r}
employment_data <- data.frame(employment_names, summary_employ)
employment_data
```

Notice that the response id variable has traveled along. We need to remove it.

```{r}
employment_data <- employment_data[2:10,]
employment_data
```

#3 Graph the results

```{r, out.width = "500px"}
ggplot(data = employment_data, aes(x = employment_names, y = summary_employ )) +
  geom_bar(stat="identity") + 
  scale_x_discrete(guide = guide_axis(angle = 45))
ggplot(data = employment, aes(x = `Employment_Employed, full-time`)) +
  geom_bar()
```

## How to analyze results across two questions.

Suppose we wanted to know which languages have been used by people that are employed full time. 

#1 Read in both sets of data

```{r}
question1 <- read_csv("https://javastats.com/data/asa_data_challenge/survey_results_expand_on_column_Employment.csv")

question2 <- read_csv("https://javastats.com/data/asa_data_challenge/survey_results_expand_on_column_LanguageHaveWorkedWith.csv")
```


#2 Join these data sets together

```{r}
combined_data <- left_join(question1, question2)
```

#3 Subset the data by variable of interest

We want to focus only on the people that are employed fulltime.

```{r}
full_time <- combined_data %>% filter(`Employment_Employed, full-time` == 1)
# Notice that the column name is surrounded by back ticks, this is necessary
# when the variable name has spaces in it
```

Now we can go through the process we had above for summarizing and graphing the data.

```{r}
# Find the column sums
full_time_summary <- colSums(full_time)

# Get the outcome names
full_time_names <- names(full_time_summary)

# Create a data frame to store both vectors
full_time_coding <- data.frame(full_time_names, full_time_summary)

# Remove the columns that are not of interest
full_time_coding <- full_time_coding[11:61,]
```

For this example, we have 61 different options for programming languages. That is way too many. We can reduce this by focusing on only the most popular outcomes.

```{r}
full_time_coding %>% arrange(desc(full_time_summary))
```

The top 10 langauges have more than 10000 responders. We can subset the data to only focus on these languages.

```{r}

full_time_coding_top10 <- full_time_coding %>% filter(full_time_summary > 10000)

```

#4 Graphing the data

```{r}
plot <- ggplot( data = full_time_coding_top10, aes(x=full_time_names, y = full_time_summary)) +
          geom_bar(stat="identity") + scale_x_discrete(guide = guide_axis(angle = 45))

plot_data <- combined_data %>% 
         select(`Employment_Employed, full-time`, LanguageHaveWorkedWith_JavaScript) %>%
         mutate(FullTime = as.factor(`Employment_Employed, full-time`), 
                JavaScript = as.factor(LanguageHaveWorkedWith_JavaScript)) %>%
         select(FullTime, JavaScript)
ggplot(data = plot_data, aes(x = FullTime, fill = JavaScript)) + geom_bar()


plot_data_programming <- combined_data %>% 
         select(`Employment_Employed, full-time`, LanguageHaveWorkedWith_JavaScript, LanguageHaveWorkedWith_SQL) %>%
         mutate(FullTime = as.factor(`Employment_Employed, full-time`), 
                JavaScript = as.factor(LanguageHaveWorkedWith_JavaScript),
                SQL = as.factor(LanguageHaveWorkedWith_SQL)) %>%
         select(FullTime, JavaScript, SQL)
ggplot(data = plot_data_programming, aes(x = FullTime, fill = JavaScript)) + 
  geom_bar() + 
  facet_grid(~SQL)
```

#5 Clean up the labels.

```{r}
# Convert the summary column to a factor

full_time_coding_top10$full_time_names <- as.factor(full_time_coding_top10$full_time_names)

# Create the labels we want
responses <- c("all shells", "C#", "C++", "HTML/CSS", "Java", "Javascript", "PHP", "Python", "SQL", "Typescript")

# Attache the labels
levels(full_time_coding_top10$full_time_names) <- responses

# Graph again
ggplot( data = full_time_coding_top10, aes(x=full_time_names, y = full_time_summary)) +
          geom_bar(stat="identity") + scale_x_discrete(guide = guide_axis(angle = 45)) + labs(title = "Languages used by people employed full time", y = "Count", x = "languages")
```

