---
title: "Exploring the Survey Data"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#read in package
library(tidyverse)
```

## Reading in our Data

We'll read in our data from a URL using the `read_csv()` function.

```{r}
survey <- read_csv("https://www4.stat.ncsu.edu/~online/datasets/survey_results_public.csv")
survey
```


## Manipulating Data

As much of the data has been read in as character data, we have a good bit of work to do getting it ready to plot and summarize!

Let's look at a manipulation of two variables: `EdLevel` and `AISent`.

First, what values do the two variables take on?

```{r}
#look at the values the EdLevel variable can take on
survey %>%
  select(EdLevel) %>%
  unique()

survey %>%
  select(AISent) %>%
  unique()
```

Now we can create new variables that are `factor` versions of these variables. As there is an inherent ordering, we'll also add in `ordered = TRUE`. This will make plots put the levels in a certain order!

```{r}
#add a new 'factor' version of the variable to the data set
survey <- survey %>% 
  mutate(EdLevel2 = factor(EdLevel, 
                           levels = c(
                             "Something else", 
                             "Primary/elementary school",                                                         
                             "Secondary school (e.g. American high school, German Realschule or Gymnasium, etc.)",
                             "Some college/university study without earning a degree",
                             "Associate degree (A.A., A.S., etc.)",                                               
                             "Bachelor’s degree (B.A., B.S., B.Eng., etc.)",                                      
                             "Master’s degree (M.A., M.S., M.Eng., MBA, etc.)",                                   
                             "Professional degree (JD, MD, Ph.D, Ed.D, etc.)"                                    
                           ), 
                           labels = c("Other", 
                                      "Primary/Elementary", 
                                      "High School", 
                                      "Some College",
                                      "Associate's",
                                      "Bachelor's",
                                      "Master's", 
                                      "Professional Degree"),
                           ordered = TRUE),
         AISentiment = factor(AISent, 
                              levels = c("Very unfavorable", 
                                         "Unfavorable", 
                                         "Indifferent", 
                                         "Very favorable", 
                                         "Favorable", 
                                         "Unsure"),
                              ordered = TRUE)
         )
```

## Plotting

Now we can do some basic plots with these variables. To remove the `NA` (or missing values), we'll first drop that from the data with `drop_na()`.

```{r}
ggplot(survey %>% 
         select(EdLevel2) %>% 
         drop_na(), 
       aes(x = EdLevel2)) + 
  geom_bar()
```

Add in some levels to pretty things up.

```{r}
ggplot(survey %>% 
         select(EdLevel2) %>% 
         drop_na(), 
       aes(x = EdLevel2)) + 
  geom_bar() +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  labs(x = "Education Level",
       title = "Most Respondents have a Bachelor's or Master's Degree") 
```

We can involve a second variable by filling in the bars by another variable (here `AISentiment`).

```{r}
ggplot(survey %>% 
         select(EdLevel2, AISentiment) %>% 
         filter(AISentiment != "Unsure") %>%
         drop_na(), 
       aes(x = EdLevel2, fill = AISentiment)) + 
  geom_bar() +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  labs(x = "Education Level",
       title = "Most Respondents have a Bachelor's or Master's Degree") 
```

We can bring a third variable in by faceting the plot. This creates the same plot at each value of the faceting variable. This can be done with `facet_wrap()`. We'll consider the `SOAccount` variable which has three values it takes on. That means we'll get three plots out!

```{r}
ggplot(survey %>% 
         select(EdLevel2, AISentiment, SOAccount) %>% 
         filter(AISentiment != "Unsure") %>%
         drop_na(), 
       aes(x = EdLevel2, fill = AISentiment)) +
  geom_bar() +
  facet_wrap(~SOAccount) + 
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  labs(x = "Education Level",
       title = "Most Respondents have a Bachelor's or Master's Degree", 
       subtitle = "Favorable or Very Favorable Opinions of AI are Consistent Across Education Groups\n
Trends are similar for those with and without a 
Stack Overflow Account")
```

